---
title: "coalesce"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coalesce}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<code>replyr::coalesce()</code> is used to fill in missing data especially for 
data coming from counting aggregations.  The issue is that for data representing
counts you often do want rows for zero-counts, and data produced by un-weighted
counting can never produce such rows (please see ["The Zero Bug"](http://www.win-vector.com/blog/2017/02/the-zero-bug/) for some discussion).

<code>replyr::coalesce()</code> takes two primary arguments  `data` (the data to be augmented) and
`support` (the pattern of rows defining the complete desired range of the data).  `support`
is taken as a list of keys.  If `support` can not be unique joined into `data` (say `data` has
rows that are do not correspond to `support` key combinations) then <code>replyr::coalesce()</code>
intentionally throws to avoid silent data loss.  Otherwise an anti-join is performed to discover
new rows that need to be added to `data` and column in this new row set are populated either by `NA`
or by scalar values from the named-list argument `fills`.  <code>replyr::coalesce()</code> works
over various `dplyr` controlled data services (Spark 2 and above, PostgreSQL, SQLite, and local data).

Some examples are given below.

Setting up:

```{r setup}
library("dplyr")
library("replyr")
```

A simple single key example:


```{r simpleexample}
# single column key example
data <- data.frame(year = c(2005,2007,2010),
                   count = c(6,1,NA),
                   name = c('a','b','c'),
                   stringsAsFactors = FALSE)
support <- data.frame(year=2005:2010)
filled <- replyr_coalesce(data, support,
                          fills=list(count=0))
filled <- filled[order(filled$year), ]
filled
```

A cross product of keys example:

```{r crossproductexample}
# complex key example
data <- data.frame(year = c(2005,2007,2010),
                   count = c(6,1,NA),
                   name = c('a','b','c'),
                   stringsAsFactors = FALSE)
support <- expand.grid(year=2005:2010,
                   name= c('a','b','c','d'),
                   stringsAsFactors = FALSE)
filled <- replyr_coalesce(data, support,
                          fills=list(count=0))
filled <- filled[order(filled$year, filled$name), ]
filled
```

An irregular (cities contained in counties) example:

```{r irregularexample}
# cities and counties
def <- data.frame(county= c('Calaveras County',
                            'Colusa County',
                            'Colusa County'),
                  city= c('Angels Camp',
                          'Colusa',
                          'Williams'),
                  stringsAsFactors= FALSE)
counts <- data.frame(county= c('Calaveras County',
                               'Colusa County'),
                     city= c('Angels Camp',
                             'Colusa'),
                     n= c(2,3),
                     stringsAsFactors= FALSE)
replyr::replyr_coalesce(counts, def, fills = list(n=0))
```

Re-running and example in a database:

```{r checkpkg}
execute_vignette <- requireNamespace("RSQLite", quietly = TRUE)
```

```{r dbexample, eval=execute_vignette}

my_db <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
RSQLite::initExtension(my_db)
dbData <- dplyr::copy_to(my_db, data)
dbSupport <- dplyr::copy_to(my_db, support)
replyr::replyr_coalesce(dbData, dbSupport, 
                        fills = list(count=0))
```

As we mentioned <code>replyr_coalesce</code> only alter column that it has added.  Under this convention the column "<code>n</code>" is not zeroed in the "<code>c</code>-row".

```{r fill, eval=execute_vignette}
d <- data.frame(x=c('a','c'), n=c(1,NA), 
                stringsAsFactors=FALSE)
s <- data.frame(x=c('a','b','c'), 
                stringsAsFactors=FALSE)

print(d)

replyr::replyr_coalesce(d,s,
                        fills= list(n=0))
```

Also <code>replyr_coalesce</code> intentionally throws if the join discovers new rows in `data`.  This is to help defend against silent data loss.

```{r detect, eval=execute_vignette}
d <- data.frame(x=c('a','c'), n=c(1,NA), 
                stringsAsFactors=FALSE)
s <- data.frame(x=c('a','b'), 
                stringsAsFactors=FALSE)
tryCatch(
  replyr::replyr_coalesce(d,s,
                        fills= list(n=0)),
  error= function(e) { e })
```

Cleaning up:

```{r cleanup}
DBI::dbDisconnect(my_db)
rm(list=ls())
gc()
```


