---
title: "let Example"
author: "Nina Zumel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{let Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # ",
  fig.width = 7
)
options(width =100)
```

An example of using `let` to wrap `dplyr` expressions as functions.

```{r setup, warning=FALSE, message=FALSE}
library("dplyr")
library("replyr")
```

The desired task: write a function that takes a data frame with a specified
numerical column and an optional grouping column, and returns a data frame with 
one row per group containing:

* the mean value
* the upper and lower bounds of a +/- 1 standard deviation around the mean
* the median value
* the upper and lower bounds of an interval +/- one-half the IQR around the median.

The `dplyr` expression for such a table is easy when the column names
are known, but complicated when they are not. We use `wrapr::let` to write such a
function without the use of `lazyeval` or `rlang`/`tidyeval`.

```{r sumstat_intervals}
sumstat_intervals = function(dframe, colname, groupcolname = NULL) {
  mapping = list(COLNAME  =colname,
                 GROUPCOLNAME = groupcolname)
  let(alias = mapping,
      {
        if(!is.null(groupcolname)) {
          dframe <- group_by(dframe, GROUPCOLNAME)
        }
        summarize(dframe, 
                  sdlower = mean(COLNAME)-sd(COLNAME),
                  mean = mean(COLNAME),
                  sdupper = mean(COLNAME) + sd(COLNAME),
                  iqrlower = median(COLNAME)-0.5*IQR(COLNAME),
                  median = median(COLNAME),
                  iqrupper = median(COLNAME)+0.5*IQR(COLNAME))
      })
}
```

We can test `sumstat_intervals` on `iris`:

```{r iris1}
sumstat_intervals(iris, "Sepal.Length")
```

```{r iris2}
sumstat_intervals(iris, "Sepal.Length", "Species")
```

```{r iris3}
sumstat_intervals(iris, "Petal.Length", "Species")
```

