---
title: "let Example"
author: "Nina Zumel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{let Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # ",
  fig.width = 7
)
options(width =100)
```

An example of using `let` to wrap `dplyr` expressions as functions.

```{r setup, warning=FALSE, message=FALSE}
library("dplyr")
library("replyr")
```

The desired task: write a function that takes a data frame with a specified
numerical column and an optional grouping column, and returns a data frame with 
one row per group containing:

* the mean value
* the upper and lower bounds of a +/- 1 standard deviation around the mean
* the median value
* the upper and lower bounds of an interval +/- one-half the IQR around the median.

The `dplyr` expression for such a table is easy when the column names
are known, but complicated when they are not. We use `wrapr::let` to write such a
function without the use of `lazyeval` or `rlang`/`tidyeval`.

```{r sumstat_intervals}
sumstat_intervals = function(dframe, colname, groupcolname = NULL) {
  mapping = list(col=colname)
  if(!is.null(groupcolname)) {
    dframe <- group_by_(dframe, groupcolname)
  }
  let(alias = mapping,
      expr={
        summarize(dframe, 
                  sdlower = mean(col)-sd(col),
                  mean = mean(col),
                  sdupper = mean(col) + sd(col),
                  iqrlower = median(col)-0.5*IQR(col),
                  median = median(col),
                  iqrupper = median(col)+0.5*IQR(col))
      })
}
```

We can test `sumstat_intervals` on `iris`:

```{r iris1}
sumstat_intervals(iris, "Sepal.Length")
```

```{r iris2}
sumstat_intervals(iris, "Sepal.Length", "Species")
```

```{r iris3}
sumstat_intervals(iris, "Petal.Length", "Species")
```

We can also use `let` to parameterize other functions that specify their parameters via non-standard evaluation. For example, we could write a `ggplot2` function to plot the information in `sumstat_intervals` (either the mean-centered interval or the median-centered one) using `ggplot2::aes_string`. Or we could use `wrapr::let`.

```{r ggplotf}
plot_distributions = NULL

if  (requireNamespace("ggplot2")) {
  library("ggplot2")
  plot_distributions = function(dframe, colname, groupcol,
                                intervaltype="mean", title="") {
    if(!(intervaltype %in% c("mean", "median")))
      error("Intervaltype must be one of 'mean' or 'median'")
    
    sintervals = sumstat_intervals(dframe, colname, groupcol)
    
    # I could do the following with aes_string, but what the heck
    mapping = list(xval=groupcol, yval=colname, center=intervaltype)
    if(intervaltype=="mean") {
      mapping2 =list(lower="sdlower", upper="sdupper")
    } else {
      mapping2 =list(lower="iqrlower", upper="iqrupper")
    }
    mapping = c(mapping, mapping2)
    
    let(alias=mapping,
        expr = {
          ggplot(dframe, aes(x=xval,color=xval)) +
            geom_jitter(aes(y=yval), width=0.2, height=0, alpha=0.5) +
            geom_crossbar(data=sintervals, aes(y=center, ymin=lower, ymax=upper)) +
            ggtitle(title) + theme(plot.title=element_text(hjust=0.5)) +
            scale_color_brewer(palette="Dark2")
        })
  }
}
```

```{r ggplot1}
if(!("NULL") %in% class(plot_distributions)) {
  plot_distributions(iris, "Sepal.Length", "Species",
                     title="Iris sepal length with mean +/1 one standard deviation")
}
```

```{r ggplot2}
if(!("NULL") %in% class(plot_distributions)) {
  plot_distributions(iris, "Petal.Width", "Species",
                     intervaltype="median",
                     title="Iris petal width with median and centered IQR interval")
}
```
