---
title: "NA rm"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NA rm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width =100)
```



Removing `NA`s with `dplyr`.

## Setup

```{r start}
library('dplyr')
library('replyr')
d <- data.frame(x=c(1,2,2),y=c(3,5,NA),z=c(NA,'a','b'),
                stringsAsFactors = FALSE)
print(d)

if (requireNamespace("RSQLite")) {
  my_db <- dplyr::src_sqlite(":memory:", create = TRUE)
  # my_db <- sparklyr::spark_connect(version='2.0.0', master = "local")
  class(my_db)
  dRemote <- replyr::replyr_copy_to(my_db,d,'d',rowNumberColumn='rowNum')
} else {
  dRemote <- d # local stand in when we can't make remote
}
print(dRemote)
```

`na.omit` and `complete.cases` are the usual ways to detect and eliminate `NA`s in a local `data.frame` (Note: we don't consider removal a correct solution when building predictive models, see please [here](https://arxiv.org/abs/1611.09477) for some discussion on other data cleaning strategies.)  However, they do not work on remote `tbl` types:

### complete.cases

```{r completecases, error=TRUE}
complete.cases(d)

complete.cases(dRemote)
```

You can't rely on "stack overflow" solutions that look "dplyr-y" to work on remote data.
For example: http://stackoverflow.com/questions/22353633/filter-for-complete-cases-in-data-frame-using-dplyr-case-wise-deletion

```{r cc2, error=TRUE}
d %>% filter(complete.cases(.))

dRemote %>% filter(complete.cases(.))
```


The solution is to map `NA` managing functions into the `dplyr` controlled remote data item.

We would think can compose this use "pure dplyr operations" as follows, but again we 
end up with differences in local and remote performance.

```{r completecasesE1, error=TRUE}
d$rowNum <- seq_len(nrow(d))
d %>% mutate_all(funs(is.na)) %>%
  mutate(nNAinRow=rowSums(.)-rowNum)

dRemote %>% mutate_all(funs(is.na)) %>%
  mutate(nNAinRow=rowSums(.)-rowNum)
```

What we see is that "what is pure `dplyr`" depends on what operators are available on the service providing back-end.  For example the `mutate_if` forms (equivilent to `SQL` `UPDATE WHERE`) are currently not available for remote sources:

```{r mutateif, error=TRUE}
dRemote %>% mutate_if(TRUE,is.na) # not correct code, just to trigger "local sources" msg
```

At some point we admit we are going to have to work over the columns by hand (though we will try to keep them remote). 

```{r countnacolumns}
dRemote %>% mutate(nna=0) %>%
  mutate(nna=nna+ifelse(is.na(x),1,0)) %>% 
  mutate(nna=nna+ifelse(is.na(y),1,0)) %>% 
  mutate(nna=nna+ifelse(is.na(z),1,0))  
```

Because remote data sources may not have the same value coercion rules as `R` we
should get used to being more careful about types (hence the `ifelse`).

This means we have to parameterize variable names to write code like the following:

```{r countforloop}
cols = setdiff(colnames(dRemote),'rowNum')
dRemote %>% mutate(nna=0) -> dTmp
for(ci in cols) {
  dTmp %>% 
    mutate_(.dots=stats::setNames(paste0('nna+ifelse(is.na(',ci,'),1,0)'),'nna')) -> 
    dTmp
}
print(dTmp)
```

Or using `replyr::let`:

```{r countforloopr}
cols = setdiff(colnames(dRemote),'rowNum')
dRemote %>% mutate(nna=0) -> dTmp
for(ci in cols) {
  let(list(TARGETCOL=ci),
      dTmp %>% mutate(nna=nna+ifelse(is.na(TARGETCOL),1,0)) -> dTmp
  )
}
print(dTmp)
```


What we see is that "what is pure `dplyr`" depends on what operators are available on the service providing back-end.


### na.omit

```{r naomit, error=TRUE}
na.omit(d)

no.omit(dRemote)
```

It is fairly clear how we can build `na.omit` from `complete.cases`.

## Cleanup

```{r cleanup}
rm(list=ls())
gc()
```

