---
title: "Parametric Programming in R"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric Programming in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # "
)
options(width =100)
```

Consider the problem of "parametric programming."  That is: simply writing correct
code before knowing some details, such as the names of the columns your
procedure will have to be applied to in the future.

Suppose, for example, your task was to and build a new advisory column that tells
you which values in a column of a `data.frame` are missing or `NA`.  We will illustrate this in [R](https://cran.r-project.org) using the example data given below:

```{r}
d <- data.frame(x = c(1, NA))
print(d)
```

Performing an ad hoc analysis is trivial in `R`: we would just
directly write:

```{r eval=FALSE}
d$x_isNA <- is.na(d$x)
```

We used the fact that we are looking at the data interactively to note the only column is "`x`", and
then picked "`x_isNA`" as our result name.  If we want to use  [`dplyr`](https://CRAN.R-project.org/package=dplyr )
the notation remains straightforward:

```{r}
library("dplyr")
packageVersion("dplyr")
d %>% mutate(x_isNA = is.na(x))
```

Now suppose, as is common in actual data science and data wrangling work, we
are not the ones picking the column names.  Instead suppose we are trying to produce
reusable code to perform this task again and again on many data sets.  In that
case we would then expect the column names to be given to us as values inside
other variables (i.e., as parameters).

```{r}
cname <- "x"                            # column we are examining
rname <- paste(cname, "isNA", sep= '_') # where to land results
print(rname)
```

And writing the matching code is again trivial:

```{r, eval=FALSE}
d[[rname]] <- is.na(d[[cname]])
```

We are now programming at a slightly higher level, or automating tasks.  We don't need
to type in new code each time a new data set with a different column name comes in.  It
is now easy to write a `for-loop` or `lapply` over a list of columns to analyze many columns
in a single data set.  It is an absolute travesty when something that is purely virtual (such 
as formulas and data) can not be automated over.  So the slightly clunkier "`[[]]`" notation (which
can be automated) is a necessary complement to the more convenient "`$`" notation (which is too specific
to be easily automated over).

Using `dplyr`
directly (when you know all the names) 
is deliberately straightforward, but programming over `dplyr` (as of May 12, 2017, prior to `dplyr` `0.6*` and the conversion to [`rlang`/`tidyeval`]( https://CRAN.R-project.org/package=rlang) interfaces) can  become a challenge.

## Standard practice

The standard parametric `dplyr` practice is to use `dplyr::mutate_` (the standard 
evaluation or parametric variation of `dplyr::mutate`). Unfortunately 
the notation in using such an "underbar form" is currently cumbersome.

You have the choice building up your formula through variations of
one of:

 * A formula
 * Using `quote()`
 * A string
 
(source: `dplyr` Non-standard evaluation vignette "nse", for additional theory and upcoming official solutions please see [here](https://rpubs.com/hadley/lazyeval)).

Let us try a few of these to try and emphasize we are proposing a new solution, not because
we do not know of the current solutions, but instead because we are familiar with the current solutions.

### Formula interface

Formula interface is a nice option as it is `R`'s common way for holding names
unevaluated.  The code looks like the following (using `lazyeval::interp` to execute):
 
```{r}
if  (requireNamespace("lazyeval")) {
  print(d %>%
    mutate_(.dots = stats::setNames(list(
      lazyeval::interp(~ is.na(VAR),
                       VAR = as.name(cname))), rname)))
}
```

The extra layer of `list()`-wrapping seems to be needed to successfully control
both the name of the input and result (please see [here]( http://www.win-vector.com/blog/2017/02/my-recent-barug-talk-parametric-programming-in-r-with-replyr/#comment-66488) for the original solution).

Currently `mutate_` does not take "two-sided formulas" so we can not write:


```{r, error=TRUE}
if  (requireNamespace("lazyeval")) {
  print(d %>% mutate_(RCOL = lazyeval::interp(RES ~ is.na(VAR),
                                              VAR= as.name(cname),
                                              RES= as.name(rname))))
}
```


### Trying `quote()` / `substitute()`

`quote()` can delay evaluation, but isn't the right tool for parameterizing
(what the linked NSE reference called "mixing constants and variable").  We can
use the related `substitute()` method as shown below (notice `mutate_` ingores
our first attempt to name the result and we have to guess/reconstruct what
name `mutate_` ends up using).

```{r}
d %>% mutate_(.dots =
    stats::setNames(list(substitute(is.na(XVAR),list(XVAR=cname))),
                    rname))
```

My point is: even if this is something that *you* know how to accomplish, 
this is evidence we are really trying to swim upstream with this notation.

### String solutions 

String based solutions can involve using `paste` to get parameter values
into the strings.  Here is an example:

```{r}
# dplyr mutate_ paste stats::setNames solution
d %>% mutate_(.dots =
                stats::setNames(paste0('is.na(', cname, ')'),
                rname))
```

Or just using strings as an interface to control `lazyeval::interp` (without the formula interface):

```{r}
# dplyr mutate_ lazyeval::interp solution
if  (requireNamespace("lazyeval")) {
  print(d %>% mutate_(.dots =
                        stats::setNames(list(lazyeval::interp("is.na(cname)",
                                         cname = as.name(cname))), rname)))
}
```

## Our advice

Our advice is to give [`wrapr::let`](https://github.com/WinVector/replyr) a try.
`wrapr::let` takes a name mapping list (called "`alias`") and a code-block (called "`expr`").
The code-block is re-written so that names in `expr` appearing on the left hand sides
of the `alias` map are replaced with names appearing on the right hand side of the `alias` map.

The code looks like this:

```{r}
# wrapr::let solution
wrapr::let(alias = list(cname = cname, rname = rname),
            expr  = {
            d %>% mutate(rname = is.na(cname))
            })
```

Notice we are able to use `dplyr::mutate` instead of needing to invoke `dplyr::mutate_`.
The expression block can be arbitrarily long and contain deep pipelines.  We now have
a useful separation of concerns, the mapping code is a wrapper completely outside of the user
pipeline (the two are no longer commingled). For complicated
tasks the ratio of `wrapr::let` boilerplate to actual useful work goes down quickly.

We also have a varation for piping into (though to save such pipes for later you use 
`wrapr::let`, not `replyr::letp`):

```{r}
# replyr::letp solution
d %>% replyr::letp(alias = list(cname = cname, rname = rname),
                   expr  = {
                   . %>% mutate(rname = is.na(cname))
                   })
```

The alias map is deliberately only allowed to be a string to string map (no environments, `as.name`, `formula`, expressions, or values) so `wrapr::let` *itself* is easy to use in automation or program over. 
I'll repeat that for emphasis: externally `wrapr::let` is completely controllable through standard
(or parametric) evaluation interfaces.
Also notice the code we wrote is never directly mentions "`x`" or "`x_isNA`" as it pulls these names out of its execution environment.

All of these solutions have consequences and corner cases.  Our (biased) opinion is: we dislike `wrapr::let` the least.

## More reading

Our group has been writing *a lot* on `wrapr::let`.  It is new code, yet something we think analysts should try.  Some of our recent notes include:

 * [The original proposal](http://www.win-vector.com/blog/2016/12/parametric-variable-names-and-dplyr/)
 * [A non-trivial example](http://www.win-vector.com/blog/2016/12/using-replyrlet-to-parameterize-dplyr-expressions/)
 * [The `wrapr::let` help documentation](http://www.win-vector.com/blog/2016/12/helplet-packagereplyr/)
 * [The `replyr` package README](https://github.com/WinVector/replyr/blob/master/README.md).
