---
output:
  md_document:
    variant: markdown_github
---

Check durability of `dplyr::arrange` through `dplyr::compute`.

<!-- Generated from .Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # "
)
options(width =100)
```


```{r ready}
library('dplyr')
library('RPostgreSQL')
packageVersion('dplyr')
packageVersion('RPostgreSQL')
my_db <- dplyr::src_postgres(host = 'localhost',port = 5432,user = 'postgres',password = 'pg')
class(my_db)
set.seed(32525)
dz <- dplyr::copy_to(my_db,data.frame(x=runif(1000)),'dz99',overwrite=TRUE)
```

Notice below: no warnings in frame or runtime.

```{r direct}
dz %>% arrange(x) %>% mutate(ccol=1) %>% mutate(rank=cumsum(ccol))  -> dz1
print(dz1)
warnings()
```

Notice below: warning "Warning: Windowed expression 'sum("ccol")' does not have explicit order.".  Result may appear the same, but we do not seem to be able to depend on that.

```{r compute}
dz %>% arrange(x) %>% compute() %>% mutate(ccol=1) %>% mutate(rank=cumsum(ccol))  -> dz2
print(dz2)
warnings()
```

Notice below: warning "Warning: Windowed expression 'sum("ccol")' does not have explicit order.".  Result may appear the same, but we do not seem to be able to depend on that.

```{r collapse}
dz %>% arrange(x) %>% collapse() %>% mutate(ccol=1) %>% mutate(rank=cumsum(ccol))  -> dz3
print(dz3)
warnings()
```


```{r printversion}
version
```