---
output:
  md_document:
    variant: markdown_github
---

Be aware different `dplyr` back-ends represent `NA` much differently.  Expect numeric `NA` to be presented as `NaN` quite often, and expect database based implementations to use `NULL` (in their sense of `NULL`, *not* in R's sense) especially in character types.  Also some `dplyr` back-ends may not have a currently accessible `NULL` concept for character types (such as Spark).

`dplyr` `0.5.0` with `RMySQL` `0.10.9` (both current on [Cran](https://cran.r-project.org) 11-27-2016) failing to insert `NULL` into `MySQL` (filed as [dplyr issue 2259](https://github.com/hadley/dplyr/issues/2259), status moved to duplicate of [dplyr issue 2256](https://github.com/hadley/dplyr/issues/2256)).

```{r mysqlnull, error=TRUE}
library('dplyr')
library('nycflights13')
packageVersion('dplyr')
packageVersion('RMySQL')
mysql <- src_mysql('mysql','127.0.0.1',3306,'root','passwd')
flts <- flights
flights_mysql <- copy_to(mysql,flts,
  temporary = TRUE,overwrite = TRUE,
  indexes = list(c("year", "month", "day"), "carrier", "tailnum"))
```

`Spark` `2.0.0` with `sparklyr` `0.4.26` not faithful to `NA` values in character
or factor columns of `data.frame`.  As we see below they get converted to blank
in a round trip between local `data.frame`s and `Spark` representations. Obviously
the round trip can not be fully faithful (we fully expect factors types to become character types, and can live with numeric `NA` becoming `NaN`) due to differences in representation.  But `Spark` can represent missing values in character columns (for example see [here](http://stackoverflow.com/questions/32067467/create-new-dataframe-with-empty-null-field-values)).

Filed as [sparklyr issue 340](https://github.com/rstudio/sparklyr/issues/340).

```{r sparklyr200}
library('sparklyr')
packageVersion('sparklyr')
s200 <- my_db <- sparklyr::spark_connect(version='2.0.0', 
   master = "local")

d2 <- data.frame(x=factor(c('z1',NA,'z3')),y=c(3,5,NA),z=c(NA,'a','z'),
                 stringsAsFactors = FALSE)
print(d2)

d2r <- copy_to(s200,d2,'d2',
               temporary = FALSE,overwrite = TRUE)
print(d2r)
d2x <- as.data.frame(d2r)
print(d2x)
summary(d2x)
str(d2x)
```

```{r printversion}
version
```

