---
output:
  md_document:
    variant: markdown_github
---

### Altering captured reference damages spark results.

If you use a variable in `dplyr::mutate()` against a `sparklyr` data source the lazy eval captures references to user variables.  Changing values of those variables implicitly changes the `mutate` and changes the values seen in the `sparklyr` result (which is itself a query).  This can be worked around by dropping in `dplyr::compute()` but it seems like it can produce a lot of incorrect calculations.  Below is a small example and a lot information on the versions of everything being run.  I am assuming the is a `sparklyr` issue as the query views are failrly different than a number of other `dplyr` structures, but it could be a `dplyr` issue.

<!-- Generated from .Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # "
)
options(width =100)
```


OSX 10.11.6.
Spark installed as described at http://spark.rstudio.com

```
library('sparklyr')
spark_install(version = "2.0.0")
```

```{r setup}
library('dplyr')
library('sparklyr')
R.Version()$version.string
packageVersion('dplyr')
packageVersion('sparklyr')
my_db <- sparklyr::spark_connect(version='2.0.0', master = "local")
class(my_db)
my_db$spark_home
print(my_db)
```

 * Expected outcome: `s1` has the same value 
 * Observed outcome: changing varaible v changes `s1` column.

```{r issue}
support <- copy_to(my_db,
                   data.frame(year=2005:2010),
                   'support')
v <- 0
s1 <- dplyr::mutate(support,count=v)

print(s1) # print 1

# s1 <- dplyr::compute(s1) # likely work-around
v <- ''

print(s1) # print 2
```

Notice `s1` changed its value (like due to lazy evaluation and having captured a reference to `v`).


Submitted as [sparklyr issue 503](https://github.com/rstudio/sparklyr/issues/503) and [dplyr issue 2455](https://github.com/hadley/dplyr/issues/2455).

```{r printversion}
version
```


