---
output:
  md_document:
    variant: markdown_github
---

`union_all` issue with `SQLite`.  Submitted as [dplyr issue 2270](https://github.com/hadley/dplyr/issues/2270).

<!-- Generated from .Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # "
)
options(width =100)
```

```{r unionx, error=TRUE}
suppressPackageStartupMessages(library('dplyr'))
packageVersion('dplyr')
packageVersion('dbplyr')
my_db <- dplyr::src_sqlite(":memory:", create = TRUE)
dr <- dplyr::copy_to(my_db,
                     data.frame(x=c(1,2), y=c('a','b'),
                                stringsAsFactors = FALSE),
                     'dr',
                     overwrite=TRUE)
dr <- head(dr,1)
# dr <- compute(dr)
print(dr)
print(dplyr::union_all(dr,dr))
```

Filed as [RSQLite 215](https://github.com/rstats-db/RSQLite/issues/215) and [dplyr 2858](https://github.com/tidyverse/dplyr/issues/2858).

```{r cleanup}
rm(list=ls())
gc()
```

Note calling `compute` doesn't always fix the problem in my more complicated production example.
Also `union` seems to not have the same issue as `union_all`.  It also seems like nested function calls exacerbating the issue, perhaps a reference to a necissary structure goes out of scope and allows sub-table collection too soon?  To trigger the full error in `replyr` force use of `union_all` in `replyr_bind_rows` and then try knitting `basicChecksSpark200.Rmd`.

The following now works:

```{r unionxs200, error=TRUE}
suppressPackageStartupMessages(library('dplyr'))
suppressPackageStartupMessages(library('sparklyr'))
packageVersion('dplyr')
packageVersion('dbplyr')
packageVersion('sparklyr')
my_db <- sparklyr::spark_connect(version='2.0.0', 
   master = "local")
class(my_db)
my_db$spark_home
da <- dplyr::copy_to(my_db,
                     data.frame(x=c(1,2),y=c('a','b'),
                                stringsAsFactors = FALSE),
                     'da',
                     overwrite=TRUE)
da <- head(da,1)
print(da)
db <- dplyr::copy_to(my_db,
                     data.frame(x=c(3,4),y=c('c','d'),
                                stringsAsFactors = FALSE),
                     'db',
                     overwrite=TRUE)
db <- head(db,1)
#da <- compute(da)
db <- compute(db)
print(db)
res <- dplyr::union_all(da,db)
res <- dplyr::compute(res)
print(res)
print(da)
print(db)
```

```{r cleanup2}
rm(list=ls())
gc()
```

```{r printversion}
version
```
