---
title: "Basic Checks"
author: "John Mount"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " # "
)
options(width =100)
```

Example trying most of the `replyr` functions on a few data sources.

```{r setup}
library('dplyr')
library('sparklyr')
source('CheckFns.R')
```


Check `replyr` basic opearations against a few data service providers.
-------------


Local `data.frame` example.

```{r localexample}
noopCopy <- function(df,name) {
  df
}
resBase <- runExample(noopCopy)
```

Local `tbl` example.

```{r localtblexample}
tblCopy <- function(df,name) {
  as.tbl(df)
}
resTbl <- runExample(tblCopy)
if(!listsOfSameData(resBase, resTbl)) {
  stop("tbl result differs")
}
```

`SQLite` example.

```{r SQLiteexample}
my_db <- dplyr::src_sqlite(":memory:", create = TRUE)
class(my_db)
copyToRemote <- remoteCopy(my_db)
resSQLite <- runExample(copyToRemote)
if(!listsOfSameData(resBase, resSQLite)) {
  stop("SQLite result differs")
}
rm(list=c('my_db','copyToRemote')); gc() # disconnect
```

MySQL example ("docker start mysql").  Kind of poor as at least the adapted MySql 
has a hard time with `NA`.

```{r MySQLexample, eval=FALSE}
my_db <- dplyr::src_mysql('mysql','127.0.0.1',3306,'root','passwd')
class(my_db)
copyToRemote <- remoteCopy(my_db)
resMySQL <- runExample(copyToRemote)
failures <- failingFrameIndices(resBase, resMySQL) 
retrykeys <- list()
retrykeys[[2]] <- c('x', 'z')
retrykeys[[3]] <- c('x', 'z')
retrykeys[[7]] <- c('year', 'name')
retrykeys[[8]] <- c('year', 'name')
retrykeys[[9]] <- c('year')
for(i in failures) {
  if(i<=length(retrykeys)) {
    explained <- sameData(resBase[[i]], resMySQL[[i]],
                          ingoreLeftNAs= TRUE, keySet=retrykeys[[i]])
    print(paste("MySQL result differs",i,
                " explained by left NAs: ",
                explained))
    if(!explained) {
      print("MySQL non NA differnce")
    }
  } else {
    print(paste("different result for example", i))
  }
}
rm(list=c('my_db','copyToRemote')); gc() # disconnect
```

PostgreSQL example ("docker start pg").
Commented out for now as we are having trouble re-installing `RPostgreSQL`.

```{r PostgreSQLexample, eval=FALSE}
my_db <- dplyr::src_postgres(host = 'localhost',port = 5432,user = 'postgres',password = 'pg')
class(my_db)
copyToRemote <- remoteCopy(my_db)
resPostgreSQL <- runExample(copyToRemote)
if(!listsOfSameData(resBase, resPostgreSQL)) {
  stop("PostgreSQL result differs")
}
rm(list=c('my_db','copyToRemote')); gc() # disconnect
```

Spark 2.0.0. example (lowest version of Spark we are supporting).

```{r Spark200example}
# Can't easilly override Spark version once it is up.
my_db <- sparklyr::spark_connect(version='2.0.0', 
   master = "local")
class(my_db)
my_db$spark_home
copyToRemote <- remoteCopy(my_db)
resSpark <- runExample(copyToRemote)
if(!listsOfSameData(resBase, resSpark)) {
  stop("Spark result differs")
}
rm(list=c('my_db','copyToRemote')); gc() # disconnect
```

```{r done}
print("all done")
rm(list=ls())
gc()
```
